---
title: "Introduction to Data Science"
author: "Victor Coppin"
date: "2025-06-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This file follows the course [*Introduction to Data Science by Rafael A. Irizarry*](Introduction%20to%20Data%20Science.pdf)

R was developed by statisticians as an interactive environment for data analysis   
[Click here to read the History of R (PDF)](historyofS.pdf)  
- R is free and open source
- It runs on all major platforms : Windows, macOS, UNIX/Linux
- Scripts and data objects can be shared seamlessly across platforms
- Active community : [Free resources for learning R](https://stats.stackexchange.com/questions/138/free-resources-for-learning-r) 

## Installation and packages 

### Scripts and Markdown

- **Create a new Script (.R)**: `Ctrl + Shift + N`  
- **Save a Script or R Markdown file**: `Ctrl + S`

### Packages 

To install packages, inside the Console :  

```r
install.packages("ggplot2") # install the ggplot2 library
install.packages("dslabs") # install the Harvard dslabs R package
installed.packages()[,"Packages"] #list all installed packages
install.packages("tidyverse", dependencies = TRUE) # tells R to install the other things that are necessary for the package or packages to run smoothly. 
install.packages(c("tidyverse", "dslabs")） # to install two packages at the same time
"ggplot2" %in% rownames(installed.packages()) # return TRUE if a package is already installed.
```

Note: 
"We recommend installing packages through RStudio, rather than through R, and the code provided works in both R and RStudio. Once a package has been installed, it is technically added onto R (even if you use RStudio to install it), which is why packages must be re-installed when R is updated. However, since we use R through RStudio, any packages that are installed can be used in both R and RStudio, regardless of which one was used to install the packages"^[Source: *Introduction to Data Science* by Rafael A. Irizarry]

Load a package into the R session 
Note :  
- to create a chunk code block  : `Ctrl + Alt + I`
- to run a chunk : `Ctrl + Shift + Enter`
- Hides messages (*i.e.* from library calls) : `{r, message=FALSE}`
```{r, message=FALSE}
library(dslabs)
```

### Running commands

#### shortcuts: 

- **Run an entire script**: `Ctrl+S`
- **Run a single line of script**: `Ctrl+Enter` while the cursor is pointing to that specific line.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dslabs)
data(murders)

murders %>%
  ggplot(aes(population,total, label=abb, color=region ))+
  geom_label()

```
# R Basics

## Objects

### Variables 
To define a variable we may use the assignment symbol: `<-`  


Code to solve the equation : $x^2 + x - 1 = 0$
The quadratic formula gives us the solutions :
$$
\dfrac{-b-\sqrt{b^2-4ac}}{2a} \quad \text{and} \quad \dfrac{-b+\sqrt{b^2-4ac}}{2a} 
$$

```{r}
# assigning values to the variables
a <- 1
b <- 1
c <- -1

#solving the quadratic equation
(-b +sqrt(b^2-4*a*c))/(2*a)
(-b -sqrt(b^2-4*a*c))/(2*a)

```

To see the value stored in a variable we ask R to evaluate `a` or more explicitly to show us the value with the `print()` function  
All the variables are saved in the workspace: `ls()` shows the names of the objects saved.

```{r}
a
print(a)
ls()
```





### Functions

