---
title: "Introduction to Data Science"
author: "Victor Coppin"
date: "2025-06-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This file follows the course [*Introduction to Data Science by Rafael A. Irizarry*](Introduction%20to%20Data%20Science.pdf)

R was developed by statisticians as an interactive environment for data analysis   
[Click here to read the History of R (PDF)](historyofS.pdf)  
- R is free and open source
- It runs on all major platforms : Windows, macOS, UNIX/Linux
- Scripts and data objects can be shared seamlessly across platforms
- Active community : [Free resources for learning R](https://stats.stackexchange.com/questions/138/free-resources-for-learning-r) 

## Installation and packages 
### Scripts and Markdown

- **Create a new Script (.R)**: `Ctrl + Shift + N`  
- **Save a Script or R Markdown file**: `Ctrl + S`

### Packages 

To install packages, inside the Console :  

```r
install.packages("ggplot2") # install the ggplot2 library
install.packages("dslab") # install the the Harvard dslab R package
installed.packages()[,"Packages"] #list all installed packages
"ggplot2" %in% rownames(installed.packages()) #return true if a package i already installed.
```

Note: 
"We recommend installing packages through RStudio, rather than through R, and the code provided works in both R and RStudio. Once a package has been installed, it is technically added onto R (even if you use RStudio to install it), which is why packages must be re-installed when R is updated. However, since we use R through RStudio, any packages that are installed can be used in both R and RStudio, regardless of which one was used to install the packages"^[Source: *Introduction to Data Science* by Rafael A. Irizarry]

Load a package into the R session 
Note :  
- to create a chunk code block  : `Ctrl + Alt + I`
- to run a chunk : `Ctrl + Shift + Enter`
- Hides messages (*i.e.* from library calls) : `{r, message=FALSE}`
```{r, message=FALSE}
library(dslabs)
```

