---
title: "Introduction to Data Science"
author: "Victor Coppin"
date: "2025-06-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This file follows the course [*Introduction to Data Science by Rafael A. Irizarry*](Introduction%20to%20Data%20Science.pdf)

R was developed by statisticians as an interactive environment for data analysis   
[Click here to read the History of R (PDF)](historyofS.pdf)  
- R is free and open source
- It runs on all major platforms : Windows, macOS, UNIX/Linux
- Scripts and data objects can be shared seamlessly across platforms
- Active community : [Free resources for learning R](https://stats.stackexchange.com/questions/138/free-resources-for-learning-r) 

## Installation and packages 

### Scripts and Markdown

- **Create a new Script (.R)**: `Ctrl + Shift + N`  
- **Save a Script or R Markdown file**: `Ctrl + S`

### Packages 

To install packages, inside the Console :  

```r
install.packages("ggplot2") # install the ggplot2 library
install.packages("dslabs") # install the Harvard dslabs R package
installed.packages()[,"Package"] #list all installed packages
install.packages("tidyverse", dependencies = TRUE) # tells R to install the other things that are necessary for the package or packages to run smoothly. 
install.packages(c("tidyverse", "dslabs")) # to install two packages at the same time
"ggplot2" %in% rownames(installed.packages()) # return TRUE if a package is already installed.
```

Note: 
"We recommend installing packages through RStudio, rather than through R, and the code provided works in both R and RStudio. Once a package has been installed, it is technically added onto R (even if you use RStudio to install it), which is why packages must be re-installed when R is updated. However, since we use R through RStudio, any packages that are installed can be used in both R and RStudio, regardless of which one was used to install the packages"^[Source: *Introduction to Data Science* by Rafael A. Irizarry].



Load a package into the R session 
Note :  
- to create a chunk code block  : `Ctrl + Alt + I`
- to run a chunk : `Ctrl + Shift + Enter`
- Hides messages (*i.e.* from library calls) : `{r, message=FALSE}`
```{r, message=FALSE}
library(dslabs)
```

### Running commands

#### shortcuts: 

- **Run an entire script**: `Ctrl + Shift + Enter`
- **Run a single line of script**: `Ctrl+Enter` while the cursor is pointing to that specific line.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dslabs)
data(murders)

murders %>%
  ggplot(aes(population,total, label=abb, color=region ))+
  geom_label()

```
# R Basics

## Objects

### Variables 
To define a variable we may use the assignment symbol: `<-`  


Code to solve the equation : $x^2 + x - 1 = 0$
The quadratic formula gives us the solutions :
$$
\dfrac{-b-\sqrt{b^2-4ac}}{2a} \quad \text{and} \quad \dfrac{-b+\sqrt{b^2-4ac}}{2a} 
$$

```{r}
# assigning values to the variables
a <- 1
b <- 1
c <- -1

#solving the quadratic equation
(-b +sqrt(b^2-4*a*c))/(2*a)
(-b -sqrt(b^2-4*a*c))/(2*a)

```

To see the value stored in a variable we ask R to evaluate `a` or more explicitly to show us the value with the `print()` function  
All the variables are saved in the workspace: `ls()` shows the names of the objects saved.

```{r}
a
print(a)
ls()
```


### Functions

>"R includes several predefined functions and most of the analysis pipelines we construct make extensive use of these.
There are many more prebuilt functions and even more can be added through packages. These functions do not appear in the workspace because you did not define them, but they are available for immediate use.
In general, **we need to use parentheses** to evaluate a function. If you type ls, the function is not evaluated and instead R shows you the code that defines the function. If you type ls() the function is evaluated and, as seen above, we see objects in the workspace. Unlike ls, most functions require one or more arguments. Below is an example of how we assign an object to the argument of the function log."

```{r}
log(8)
log(a) #a was already assigned to 1
```

#### Help 

You can find out what the function expects and what it does by reviewing the very useful manuals included in R. 
```r
help("log")
?log # shorthand
```

>"The help page will show you what arguments the function is expecting. For example, log needs x and base to run. However, some arguments are required and others are optional. You can determine which arguments are optional by noting in the help document that a default value is assigned with =. Defining these is optional. For example, the base of the function log defaults to base = exp(1) making log the natural log by default."

```{r}
args(log) #quick look at the argument without opening the help system
```

If no argument name is used, R assume arguments are in the order shown in the help page: 

```{r}
log(x=8,base=2)
log(base=2, x=8) # if using arguments' names, we can include them in whatever order we want. 
log(8,2)
```

#### Arithmetic and relational operators 

Arithmetic and relational operators are functions that didn't need parentheses to be evaluated. 
Note: In R, operators like `+`, `*`, or `%%` are actually functions, so you can call `help("+")` just like any function.

```r
5+3 
5-1
5*3
10/2
2**2  # Exponentiation
2^3   # Exponentiation
5%%2  # Modulo 
5%/%5 # Integer division

# Help for operators
help("+") # Give a list of all arithmetic operator not only "+"  
?">"      # Help for all relational operators

```

### Other pre-built objects


- Mathematical constants like $\pi$ and $\infty$ are represented as `pi` and `Inf` in R
- List all the available data sets : `data()`
  These data sets are objects that can be used by simply typing the name :`co2`



